---
title: "Assignment 3: Data Exploration"
author: "Camille DeSisto"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 
```{r,  message = FALSE}
getwd()
setwd("C:/ENV872/Environmental_Data_Analytics_2021/Data")
getwd()
library(tidyverse)

Neonics <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Litter <- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)
```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: The ecotoxicology of neonicotinoids on insects is critical because insect pests threaten agricultural products, potentially harming the food system. Specifically, pests could lead to shortages of certain crops and/ or surges in food prices. Insecticides can be important in preventing insect pests from harming crops. However, insecticides such as neonicotinoids could pose negative externalities for non-pest insects. For example, they could harm important pollinators such as bees. Even if the neonicotinoids do not kill beneficial insects. long-term exposure could still be damaging, especially to their behavioral patterns becuase neonicotinoids affect the nervous system. Therefore, a robust understanding of the ecotoxicology of neonicotinoids on insects has important implications for both agriculture and biodiversity conservation. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Quantitative studies of litter and woody debris in forests are important because forest litter is a key indicator of forest productivity and phenological patterns. Data on litter can be used to study feedbacks with the soil, changes in climatic patterns in a forest, and how forests respond to disturbance, among others. Woody debris is equally important in studying those dynamics. For example, woody debris is a key factor to consider when studying nutrient turnover and aboveground biomass in forest ecosystems.     

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: In the NEON network, this sampling is conducted in 40m by 40m plots within larger tower plots. Litter is collected from elevated traps whereas woody debris is collected from ground traps. Dpending on the vegetation, the placement of the traps within the plots is either targeted or random. Ground traps are sampled once per year whereas elevated traps are sampled more frequently. When sampled, mass data are collected for separate functional groups (needles, leaves, seeds, branches, etc.).
*
*
*



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics)
ncol(Neonics)
nrow(Neonics)

#there are 4623 rows and 30 columns 
```

6. Using the `summary` function on the "Effects" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
#try to use summary 
summary(Neonics$Effect)

#also try using the table function 
table(Neonics$Effect)

 
```

> Answer: The two most common effects are population and mortality. These are of particular importance because they describe the population dynamics of the insects. Knowing mortality is really useful in determining the effects of the insecticide on the animals themselves. Also, knowing their population sizes is critucal for examining how the neonicotenoids are impacting the insects. Additionally, these two effects are relatively easy to collect data about because they are count data and do not require much specialized equipment. On the other hand, an effect such as genetics would be more time- and resource-intensive. 



7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
#try with the summary function
sort(summary(Neonics$Species.Common.Name))

#try with the table and sort functions 
sort(table(Neonics$Species.Common.Name))



```

> Answer: The six most common insects are: Honey Bees, Parasitic Wasps, Buff Tailed Bumble Bees, Carniolan Honey Bees, Bumble Bees, and Italian Honeybees. These insects are of particular concern because they are pollinators and therefore critical for agricultural productivity and biodiversity. Rather than harming plants, these animals help plant populations by providing ciritical pollication services. Global declines in bees present an urgent ecological problem that demands robust scientific studies.  

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)
unique(Neonics$Conc.1..Author.)
```

> Answer: The class of Conc.1..Author is a character because these concentrations are somtimes written the "/" symbols after the number. Other values have the "~" symbol before them, and some are listed as "NR". However, R only recognizes numbers as numeric values. 



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
library(ggplot2)
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 15, lty=1) +
  scale_x_continuous() + xlab("Population Year")+ylab("Number of Studies Conducted")+ theme_classic()
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color= Test.Location), bins = 15, lty=1) +
  scale_x_continuous() + xlab("Population Year")+ylab("Number of Studies Conducted")+ theme_classic()
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test location is the lab, a location which rose to increasing popularity until it experienced a shap decline in popularit in the mid 2010s. Natural field locations were the second most population location, but the number of studies peaked slightly before 2010 before a precipitous declines. There were some, but not many, articles published where the testing was completed in an artificial field location between 2000 and the mid 2010s, and even fewer artiles published where the location was undeterminable. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar() + theme_classic()+ theme(axis.text.x = element_text(size=9, angle=45))

length(Neonics$Endpoint[Neonics$Endpoint=="LOEL"])
length(Neonics$Endpoint[Neonics$Endpoint=="NOEL"])
```

> Answer: The two most common endpoint counts are LOEL and NOEL. LOEL is the "lowest observable effect level" (the lowest concentration that produced effects that were significantly different from the control). NOEL represents "no observable effect level" (the higest level that produces effects that are not significantly different from the control). 
>

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate)

#This is a factor and not a date

Litter$collectDate <- as.Date(Litter$collectDate, format="%Y-%m-%d")

unique(Litter$collectDate)

# litter was samples on August 8th and 30th 2018
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
length(unique(Litter$plotID))
```

> Answer: 12 different plots were sampled at Niwot Ridge. The "unique" function provides a list of all the unique (distinct) values in the column whereas the summary function provides information on how many observations there are of each unique value in the column. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar() + theme_classic()+ theme(axis.text.x = element_text(size=9, angle=10)) + xlab("Functional Group")
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass))+ theme_classic()+ theme(axis.text.x = element_text(size=9, angle=10))+ylab("Dry Mass")+xlab("Funcational Group")

ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass))+ theme_classic()+ theme(axis.text.x = element_text(size=9, angle=10))+ylab("Dry Mass")+xlab("Funcational Group")


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot is more effective than the violin plot in this case because the mass distribution is such that the violin plot is visualized as horizontal and virtical lines. It is difficult so discern the data distibution with this violin plot. On the other hand, the boxplot is useful for visualization in this case because the quantiles are clear to see for many of the functional groups. 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles, mixes, and twigs/ branches tend ot have the highest biomass at these sites. 